HO unification from object language to meta language

### 2024-04-01 paper/main.tex
flip pb ot ease reading
rework 6 up to 6.2

\noindent
While the original problem should be solved by \Uo{} (by assigning $f$ to $X$)
\Ue{} is not able to solve the problem in $\mathbb{T}$:
its left and right hand sides have different rigid heads:
$\lambda x.A_x$ corresponds to the \Ho{} term
\elpiIn{lam x\ uva A [x]} 
while $f$ corresponds to
\elpiIn{con"f"}.

In order to guarantee \cref{prop:simulation} we detect
lambdas that can disappear by eta contraction (the \maybeeta set
introduced in \cref{sec:nutshell}); then 
we modify the way terms are compiled and move terms in \maybeeta
from  $\mathbb{T}$ to the \linkStore and put fresh unification variables
in their place.

\elpiIn{type link-eta tm -> tm -> baselink}

\noindent
As per \cref{inv:linklhs} the term on the left hand side
(\lhs) is a variable, and its right counterpart (\rhs) is the
term in \maybeeta and it has the following property:

\begin{invariant}[\linketa \rhs]
  The \rhs of any \linketa in \linkStore has the shape $\lambda x.t$
  and $t$ is not a lambda. 
  %where $t_x$ is a \maybeeta term and $x$ is free in $t$.
  \label{inv:link-eta-right}
\end{invariant}

\linketa are kept in the link store (\linkStore) during execution
and activated when some conditions hold on \lhs or \rhs.
Link activation is implemented by extending the \elpiIn{progress1}
predicate (defined in \cref{sec:execution}).

\subsection{Detection of \maybeeta}

When compiling a term $t$ we need to determine if any
subterm $s \in \subterm{t}$ that is of the form $\lambda x. r$,
where $x$ occurs in $r$, can be a $\eta$-expansion, i.e. if
there exists a substitution $\rho$ such that $\rho (\lambda x.r) \Eo s$.
The detection of lambda abstractions that can ``disappear''
is not as trivial as it may seems, here a few examples:
% \\
% \noindent
% \begin{minipage}{.5\linewidth}%
%   \begin{gather}
%     \lambda x. f \appsep A_x\\
%     \lambda x. f \appsep x \appsep A_x
%   \end{gather}
% \end{minipage}% This must go next to `\end{minipage}`
% \begin{minipage}{.5\linewidth}
%   \begin{gather}
%     \lambda x. \lambda y. f \appsep A_x \appsep B_{y x} \label{eq:lamlam}\\
%     \lambda x. f \appsep A_x \appsep x \label{eq:lamdisc}
%   \end{gather}
% \end{minipage}
%
\begin{center}
  \begin{tabular}{lll}
    %Term & Status & Evidence \\\hline
    $\lambda x. f \appsep (A \appsep x)$ & $\in\maybeeta$ & $\rho = \{~ A \mapsto \lambda x.x ~\}$ \\
    $\lambda x. f \appsep (A \appsep x) \appsep x$ & $\in\maybeeta$ & $\rho = \{~ A \mapsto \lambda x.a ~\}$\\
    $\lambda x. f \appsep x \appsep (A \appsep x)$ & $\not\in\maybeeta$ &\\
    $\lambda x. \lambda y. f \appsep (A \appsep x) \appsep (B \appsep y \appsep x)$ & $\in\maybeeta$ & $\rho = \{~ A \mapsto \lambda x.x,~ B \mapsto \lambda y.\lambda x.y ~\}$
  \end{tabular}
\end{center}
\vspace{4pt}

% In the examples above, the first term is a \maybeeta since $A_x$ can
% reduce to $x$ by setting $A_x = \lambda x.x$, 
% the second one is not a \maybeeta since it exists no substitution
% for $A_x$ such that $A_x$ reduces to $x$ and $x$ is not free in the subterm $f \appsep x$.
\noindent
The first two examples are easy, and show how a unification variable can expose
or erase a variable and turn the resulting term in an $\eta$-expansion or not.\\
The third example shows when a variable occurs outside the scope of a unification
variable cannot be erased and hence prevent a term from being an $\eta$-expansion.\\
The last example shows the recursive nature of the check we need to implement.
The term starts with a spine of two lambdas hence the whole term
is in \maybeeta iff the inner term $\lambda y.f\appsep (A \appsep x) \appsep (B \appsep y \appsep x)$
is in \maybeeta itself, that means it could $\eta$-contract to
$f \appsep (A \appsep x)$ making $\lambda x.f \appsep (A \appsep x)$ a potential
$\eta$-expansion.\\

@@

\noindent
The rule verifies if the \Fo{} term $\lambda t$ in input is in
\maybeeta. Let $\lambda t'$ be the compiled version of $\lambda t$, then the fresh
variable \elpiIn{A} returned as the new \Ho term as in scope all the free names
in $t'$. The critical part of this compilation is the creation of the \linketa,
which links the variable \elpiIn{A} with $t$. This link creation enforce
\cref{inv:linklhs} and \cref{inv:link-eta-right}, since \lhs is a trivially a
variable and the \rhs is a term $t$ starting with the \elpiIn{lam} constructor where
$x$ is free in $t$ otherwise $t$ would not have been detected as a \maybeeta.

\begin{corollary}
  The \rhs of any \linketa has exactly one lambda abstraction.
\end{corollary}

\begin{proof}[Proof sketch]
  By contradiction, suppose that a \linketa $l$ has $t$ as \rhs and $t = \lambda
  x.\lambda y.t'_{xy}$. Two cases are to be analyses: 1) $\lambda y.t'_{xy}$ is
  a $\maybeeta$, then, by construction, \rhs would have been replaced with a the
  $\eta$-expansion of fresh variable $v$, which is a contradiction, since
  $\lambda y.t'_{xy} \neq \lambda x.v_{x}$; 2) $\lambda y.t'_{xy}$ is not an a
  \maybeeta, then neither $t$ is, which is a contradiction since \rhs is always
  a \maybeeta by construction.
\end{proof}

@@

\linketa are meant to delay the unification of ``problematic'' terms. In the
following, we call \linkStore the list of suspended links.

In order to activate a \linketa $l$, we need to extend the \elpiIn{progress1} predicate
by adding new rules. After passing under all the \elpiIn{abs}
constructors of $l$, there are two cases making a \linketa to progress, 1) \lhs
is instantiated to a rigid term 2) \rhs is no more a \maybeeta or it is a term
which can be reduced to a term with rigid head. 
If \lhs is instantiated to a rigid term $t$, by \cref{prop:nf},
we know that $t$ does not contain any \maybeeta. Let $t'$ the right hand side,
if $t$ is a constant or a function application, then, $t'$, which by
construction has \elpiIn{lam} as head, should be an $\eta$-expansion. We are
therefore allowed to unify $\lambda x. t \appsep x$ (the $\eta$-expanded version
of $t$) with $t'$. Finally, if $t$ is a term with \elpiIn{lam} as head, then
it is not an $\eta$-expansion and therefore, $t$ can be unified with $t'$.

@@

Once a \linketa is activated, it can be removed from \linkStore,
otherwise, the link is kept for a further iteration of \elpiIn{progress}. Note
that this link progression enforce
\cref{prop:nf,inv:linklhs,inv:link-eta-right}: we never commit a term in
the \Ho substitution, since we make unification only when we know that the 
terms are no more \maybeeta, and when \lhs is no more a variable or \rhs is 
no more a \maybeeta, the link is removed from \linkStore.
% \printAll
%   {{\elpiIn{flam x\ fapp [fuva X, x] ~\Uo~fcon"f"}}}
%   {{A \Ue f}}
%   {{\mapping{X}{X'}{1}}}
%   {{\linketaM{}{A}{\lambda x.X'_{x}}}}
% $$
% \begin{array}{l}
%   \mathbb{P} = \{~ \lambda x.X \appsep x \Uo f ~\}\\
%   \mathbb{T} = \{~ A \Ue f ~\}\\
%   \mathbb{M} = \{~ \mapping{X}{B}{1} ~\}\\
%   \linkStore \;= \{~ \linketaM{}{A}{\lambda x.B_{x ~\}}}
% \end{array}
% $$

\printAlll
  {{{\lambda x.X \appsep x,f}}}
  {{{A,f}}}
  {{{X,B,1}}}
  {{{,A,\lambda x.B_x}}}

@@

In \cref{sec:basic-comp-limitations}, we have given two unification problems to
be run one after the other. In the following table we present the entry problem,
its compiled version, with the corresponding list of mapping and links.
% make test ONLY=7001 TEX=tex
\printAlll
  {{{\lambda x.\lambda y.(X\appsep y\appsep x),\lambda x.\lambda y.x},
    {\lambda x.(f\appsep (X\appsep x)\appsep x),Y}}}
  {{{A,\lambda x.\lambda y.x},
    {D,F}}}
  {{{Y,F,0},
    {X,E,1},
    {X,C,2}}}
  {{{,D,\lambda x.(f\appsep E_{x}\appsep x)},
    {,A,\lambda x.B_{x}},
    {x,B_{x},\lambda y.C_{y x}}}}

>>>>

\noindent
While $\lambda x.X \appsep x \Uo{} f$ does admit the solution
$\rho = \{~ X \mapsto f ~\}$, the corresponding problem in
$\mathbb{T}$ does not:
\elpiIn{lam x\ uva A [x]} and
\elpiIn{con"f"} start with different, rigid, term constructors hence
\Ue{} fails.

In order to guarantee \cref{prop:simulation} we detect
lambdas that can disappear by eta contraction (\cref{sec:etadetection}) and
we modify the term compiler put fresh unification variables
in their place: the problematic term is moved from 
from  $\mathbb{T}$ to \linkStore (\cref{sec:etacomp}). The compilation
of the problem $\mathbb{P}$ above is refined to: 
%
\printAlll
  {{{\lambda x.X \appsep x,f}}}
  {{{A,f}}}
  {{{X,B,1}}}
  {{{,A,\lambda x.B_x}}}

\noindent
As per \cref{inv:linklhs} the term on the left is a variable, and its
right counterpart is the
term in \maybeeta, and that term has the following property:

\begin{invariant}[\linketa \rhs]
  The \rhs of any \linketa %in \linkStore 
  has the shape $\lambda x.t$
  and $t$ is not a lambda. 
  %where $t_x$ is a \maybeeta term and $x$ is free in $t$.
  \label{inv:link-eta-right}
\end{invariant}

\linketa are kept in the link store \linkStore during execution
and activated when some conditions hold on \lhs or \rhs.
Link activation is implemented by extending the \elpiIn{progress1}
predicate (defined in \cref{sec:execution}).

\subsection{Detection of \maybeeta}\label{sec:etadetection}

When compiling a term $t$ we need to determine if any
subterm $s \in \subterm{t}$ that is of the form $\lambda x. r$,
where $x$ occurs in $r$, can be a $\eta$-expansion, i.e. if
there exists a substitution $\rho$ such that $\rho (\lambda x.r) \Eo s$.
The detection of lambda abstractions that can ``disappear''
is not as trivial as it may seems, here a few examples:
%
\begin{center}
  \begin{tabular}{lll}
    %Term & Status & Evidence \\\hline
    $\lambda x. f \appsep (A \appsep x)$ & $\in\maybeeta$ & $\rho = \{~ A \mapsto \lambda x.x ~\}$ \\
    $\lambda x. f \appsep (A \appsep x) \appsep x$ & $\in\maybeeta$ & $\rho = \{~ A \mapsto \lambda x.a ~\}$\\
    $\lambda x. f \appsep x \appsep (A \appsep x)$ & $\not\in\maybeeta$ &\\
    $\lambda x. \lambda y. f \appsep (A \appsep x) \appsep (B \appsep y \appsep x)$ & $\in\maybeeta$ & $\rho = \{~ A \mapsto \lambda x.x,~ B \mapsto \lambda y.\lambda x.y ~\}$
  \end{tabular}
\end{center}
\vspace{4pt}

% In the examples above, the first term is a \maybeeta since $A_x$ can
% reduce to $x$ by setting $A_x = \lambda x.x$, 
% the second one is not a \maybeeta since it exists no substitution
% for $A_x$ such that $A_x$ reduces to $x$ and $x$ is not free in the subterm $f \appsep x$.
\noindent
The first two examples are easy, and show how a unification variable can expose
or erase a variable and turn the resulting term in an $\eta$-expansion or not.\\
The third example shows that when a variable occurs outside the scope of a unification
variable it cannot be erased and can hence prevent a term from being an $\eta$-expansion.\\
The last example shows the recursive nature of the check we need to implement.
The term starts with a spine of two lambdas hence the whole term
is in \maybeeta iff the inner term $\lambda y.f\appsep (A \appsep x) \appsep (B \appsep y \appsep x)$
is in \maybeeta itself. If it is, it could $\eta$-contract to
$f \appsep (A \appsep x)$ making $\lambda x.f \appsep (A \appsep x)$ a potential
$\eta$-expansion.\\

@@

\noindent
The rule triggers when the input term \elpiIn{flam F} is in
\maybeeta. It compiles it to \elpiIn{lam F1} but puts the fresh
variable \elpiIn{A} in its place. The variable sees all the names free in
\elpiIn{lam F1}. The critical part of this rule is the creation of the \linketa,
which relates the variable \elpiIn{A} with \elpiIn{lam F1}.
This link clearly validates \cref{inv:linklhs}.

\begin{corollary}
  The \rhs of any \linketa has exactly one lambda abstraction, hence
  the rule above respects \cref{inv:link-eta-right}.
\end{corollary}

\begin{proof}[Proof sketch]
  By contradiction, suppose that the rule above triggered and that
  the \rhs of the link is $\lambda x.\lambda y.t_{xy}$.
  If $\testmaybeeta{}~\lambda y.t_{xy}$ holds the recursive call to
  \elpiIn{comp} (made by \elpiIn{comp-lam}) must have put a fresh variable
  in its place, so this case is impossible.
  Otherwise, if $\testmaybeeta{}~\lambda y.t_{xy}$ does not hold, also
  $\testmaybeeta{}~\lambda x.\lambda y.t_{xy}$ does not hold, contradicting
  the assumption that the rule triggered.
\end{proof}

@@

\linketa are meant to delay the unification of ``problematic'' terms until
we know for sure if the head lambda has to be $\eta$-contracted or not.

We extend the \elpiIn{progress1} predicate
by adding new rules. After passing under all the \elpiIn{abs}
constructors of $l$, there are two cases making a \linketa to progress, 1) \lhs
is instantiated to a rigid term 2) \rhs is no more a \maybeeta or it is a term
which can be reduced to a term with rigid head. 
If \lhs is instantiated to a rigid term $t$, by \cref{prop:nf},
we know that $t$ does not contain any \maybeeta. Let $t'$ the right hand side,
if $t$ is a constant or a function application, then, $t'$, which by
construction has \elpiIn{lam} as head, should be an $\eta$-expansion. We are
therefore allowed to unify $\lambda x. t \appsep x$ (the $\eta$-expanded version
of $t$) with $t'$. Finally, if $t$ is a term with \elpiIn{lam} as head, then
it is not an $\eta$-expansion and therefore, $t$ can be unified with $t'$.

@@

Once a \linketa is activated, it can be removed from \linkStore,
otherwise, the link is kept for a further iteration of \elpiIn{progress}. Note
that this link progression enforce
\cref{prop:nf,inv:linklhs,inv:link-eta-right}: we never commit a term in
the \Ho substitution, since we make unification only when we know that the 
terms are no more \maybeeta, and when \lhs is no more a variable or \rhs is 
no more a \maybeeta, the link is removed from \linkStore.

@@

In \cref{sec:basic-comp-limitations}, we have given two unification problems to
be run one after the other. In the following table we present the entry problem,
its compiled version, with the corresponding list of mapping and links.
% make test ONLY=7001 TEX=tex
\printAlll
  {{{\lambda x.\lambda y.(X\appsep y\appsep x),\lambda x.\lambda y.x},
    {\lambda x.(f\appsep (X\appsep x)\appsep x),Y}}}
  {{{A,\lambda x.\lambda y.x},
    {D,F}}}
  {{{X,E,1},
    {Y,F,0},
    {X,C,2}}}
  {{{,D,\lambda x.(f\appsep E_{x}\appsep x)},
    {,A,\lambda x.B_{x}},
    {x,B_{x},\lambda y.C_{y x}}}}
